---
title: "NMDS Ordination Analysis Report"
author: "Ördin v3.0 - Community Ecology Analysis Platform"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    number_sections: true
geometry: margin=1in
header-includes:
  - \usepackage{xcolor}
  - \usepackage{colortbl}
  - \usepackage{fancyhdr}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \pagestyle{fancy}
  - \fancyhead[L]{NMDS Analysis Report}
  - \fancyhead[R]{\thepage}
  - \fancyfoot[C]{Generated by Ördin v3.0}
params:
  nmds_result: NULL
  distance: "bray"
  k: 2
  permutations: 999
  stress_interp: NULL
  dataset_name: "Community Data"
  n_sites: NULL
  n_species: NULL
  trymax: 20
  autotransform: FALSE
  analysis_date: NULL
  analysis_time: NULL
  analysis_type: "NMDS Ordination"
  r_version: NULL
  vegan_version: NULL
  ordin_version: "3.0"
  platform: NULL
  os: NULL
  vegan: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      fig.align = "center", fig.pos = "H")
library(vegan)
library(knitr)
library(kableExtra)
```

\newpage

# Executive Summary

This report presents the results of **Non-metric Multidimensional Scaling (NMDS)** ordination analysis performed on `r params$dataset_name`. The analysis reduces the complexity of multivariate community data into `r params$k` dimensions while preserving rank-order dissimilarities between samples.

**Key Findings:**

- **Stress Value:** `r sprintf("%.4f", params$nmds_result$stress)` (`r params$stress_interp$grade` quality)
- **Convergence:** `r if(params$nmds_result$converged) "Successful" else "Failed"`
- **Distance Metric:** `r toupper(params$distance)` dissimilarity
- **Interpretation:** `r params$stress_interp$message`

**Recommendation:** `r params$stress_interp$recommendation`

\vspace{0.5cm}

---

\vspace{0.5cm}

# Analysis Overview

**Analysis Type:** Non-metric Multidimensional Scaling (NMDS)  
**Dataset:** `r params$dataset_name`  
**Number of Sites:** `r nrow(scores(params$nmds_result, display = "sites"))`  
**Number of Species:** `r nrow(scores(params$nmds_result, display = "species"))`  
**Distance Metric:** `r toupper(params$distance)`  
**Dimensions (k):** `r params$k`

\vspace{0.5cm}

---

\vspace{0.5cm}

# Ordination Quality Assessment

\colorbox{yellow!20}{\parbox{\dimexpr\textwidth-2\fboxsep}{
\textbf{Stress: `r sprintf("%.4f", params$nmds_result$stress)`} \quad \textbf{[Grade: `r params$stress_interp$grade`]}

\textbf{`r params$stress_interp$message`}

`r params$stress_interp$detail`

\textbf{Recommendation:} `r params$stress_interp$recommendation`
}}

\vspace{0.3cm}

**Convergence:** `r if(params$nmds_result$converged) "✓ Solution converged" else "✗ Solution did not converge"`  
**Iterations:** `r params$nmds_result$iters`  
**Non-metric R²:** `r sprintf("%.4f", 1 - params$nmds_result$stress^2)`

\vspace{0.5cm}

# NMDS Ordination Plot

```{r nmds_plot, fig.width=8, fig.height=6, dpi=300, fig.cap="NMDS ordination showing sample positions in reduced dimensional space. Points represent individual samples, with distances approximately preserving ranked dissimilarities."}
# Create NMDS plot
par(mar = c(5, 5, 4, 2))
plot(params$nmds_result, 
     type = "none", 
     main = "NMDS Ordination",
     cex.main = 1.4,
     col.main = "#2e8b57",
     cex.lab = 1.2,
     cex.axis = 1.1)

# Add sample points
points(params$nmds_result, 
       display = "sites", 
       pch = 21,
       bg = "#2e8b57", 
       cex = 2,
       col = "black",
       lwd = 1.5)

# Add sample labels (if not too many)
n_sites <- nrow(scores(params$nmds_result, display = "sites"))
if (n_sites <= 30) {
  text(params$nmds_result, 
       display = "sites", 
       cex = 0.7,
       pos = 3,
       offset = 0.5)
}

# Add stress annotation with color
stress_text <- sprintf("Stress = %.3f [%s]", 
                      params$nmds_result$stress,
                      params$stress_interp$grade)
mtext(stress_text, 
      side = 3, 
      line = 0.5, 
      col = params$stress_interp$color,
      font = 2, 
      adj = 0,
      cex = 0.9)

# Add grid for reference
grid(col = "#cccccc40", lty = 1)

# Add axes origin lines
abline(h = 0, v = 0, col = "gray60", lty = 2, lwd = 0.8)
```

\vspace{0.5cm}

# Diagnostic Plots

## Shepard Plot

The Shepard plot compares observed dissimilarities with ordination distances. Points closer to the line indicate better fit.

```{r shepard_plot, fig.width=7, fig.height=5, dpi=300, fig.cap="Shepard diagram showing the relationship between observed dissimilarities and ordination distances. The step line shows monotonic relationship, while points show actual fit."}
par(mar = c(5, 5, 4, 2))
stressplot(params$nmds_result,
          main = "Shepard Diagram",
          cex.main = 1.3,
          cex.lab = 1.1,
          pch = 16,
          cex = 0.8,
          p.col = rgb(0.18, 0.55, 0.34, 0.5),
          l.col = "red")

# Add R-squared annotation
text(x = par("usr")[2] * 0.05, 
     y = par("usr")[4] * 0.95,
     labels = sprintf("Non-metric R² = %.3f", 1 - params$nmds_result$stress^2),
     adj = c(0, 1),
     cex = 1.0,
     font = 2)
```

## Goodness of Fit by Site

```{r goodness_plot, fig.width=7, fig.height=5, dpi=300, fig.cap="Goodness of fit for each sample site. Higher values indicate better representation in the ordination space."}
par(mar = c(5, 5, 4, 2))
gof <- goodness(params$nmds_result)
site_names <- rownames(scores(params$nmds_result, display = "sites"))

barplot(sort(gof, decreasing = TRUE),
        main = "Goodness of Fit by Site",
        ylab = "Goodness of Fit",
        xlab = "Sites (ranked)",
        col = colorRampPalette(c("#ff6b6b", "#feca57", "#48dbfb", "#2e8b57"))(length(gof)),
        border = NA,
        cex.main = 1.3,
        cex.lab = 1.1,
        las = 2)

abline(h = mean(gof), col = "red", lty = 2, lwd = 2)
text(x = length(gof) * 0.7, 
     y = mean(gof) * 1.1,
     labels = sprintf("Mean = %.3f", mean(gof)),
     col = "red",
     font = 2)
```

\newpage

# Ordination Statistics

```{r stats_table}
stats_df <- data.frame(
  Statistic = c("Stress", "Quality Grade", "Convergence", "Dimensions", 
                "Distance Metric", "Iterations", "Try Max", "Non-metric R²"),
  Value = c(
    sprintf("%.4f", params$nmds_result$stress),
    params$stress_interp$grade,
    if(params$nmds_result$converged) "✓ Converged" else "✗ Not Converged",
    params$k,
    toupper(params$distance),
    params$nmds_result$iters,
    params$nmds_result$tries,
    sprintf("%.4f", 1 - params$nmds_result$stress^2)
  ),
  stringsAsFactors = FALSE
)

knitr::kable(stats_df, 
             format = "latex", 
             booktabs = TRUE,
             caption = "NMDS Analysis Statistics",
             align = c("l", "r")) %>%
  kable_styling(latex_options = c("HOLD_position", "striped"))
```

\newpage

# Site Scores

```{r site_scores}
site_scores <- as.data.frame(scores(params$nmds_result, display = "sites"))
site_scores$Sample <- rownames(site_scores)
site_scores <- site_scores[, c("Sample", colnames(site_scores)[1:(ncol(site_scores)-1)])]

knitr::kable(site_scores, 
             format = "latex", 
             booktabs = TRUE,
             digits = 4,
             caption = "NMDS Site Scores (coordinates in ordination space)",
             longtable = TRUE,
             align = c("l", rep("r", ncol(site_scores)-1))) %>%
  kable_styling(latex_options = c("HOLD_position", "repeat_header", "striped"),
                font_size = 9)
```

\vspace{0.5cm}

# Species Scores

```{r species_scores}
species_scores <- as.data.frame(scores(params$nmds_result, display = "species"))
species_scores$Species <- rownames(species_scores)
species_scores <- species_scores[, c("Species", colnames(species_scores)[1:(ncol(species_scores)-1)])]

# Calculate species importance (distance from origin)
species_scores$Importance <- sqrt(rowSums(species_scores[, -1]^2))
species_scores <- species_scores[order(species_scores$Importance, decreasing = TRUE), ]

# Show top 50 species or all if fewer
n_show <- min(50, nrow(species_scores))

knitr::kable(species_scores[1:n_show, ], 
             format = "latex", 
             booktabs = TRUE,
             digits = 4,
             caption = sprintf("NMDS Species Scores (top %d by distance from origin)", n_show),
             longtable = TRUE,
             row.names = FALSE,
             align = c("l", rep("r", ncol(species_scores)-1))) %>%
  kable_styling(latex_options = c("HOLD_position", "repeat_header", "striped"),
                font_size = 8)
```

\vspace{0.5cm}

\newpage

# Interpretation

## Ordination Quality

The NMDS analysis achieved a stress value of **`r sprintf("%.4f", params$nmds_result$stress)`**, which is classified as **`r params$stress_interp$grade`** quality according to Clarke (1993). `r params$stress_interp$detail`

`r params$stress_interp$recommendation`

The non-metric R² value of **`r sprintf("%.4f", 1 - params$nmds_result$stress^2)`** indicates that `r sprintf("%.1f%%", (1 - params$nmds_result$stress^2) * 100)` of the variance in the original dissimilarities is explained by the ordination.

## Convergence

The analysis `r if(params$nmds_result$converged) "successfully converged" else "failed to converge"` after **`r params$nmds_result$iters`** iterations (maximum: `r params$nmds_result$tries` tries). `r if(params$nmds_result$converged) "This indicates a stable solution was found." else "Consider increasing the number of dimensions (k) or trying different starting configurations."`

## Distance Metric

The **`r toupper(params$distance)`** dissimilarity index was used to quantify differences between samples. This metric is appropriate for `r if(params$distance == "bray") "abundance data and is robust to differences in sampling effort" else if(params$distance == "jaccard") "presence-absence data and focuses on shared species" else "the data type analyzed"`.

## Biological Interpretation

Samples that are close together in the ordination space have similar community composition, while samples far apart are compositionally dissimilar. The axes in NMDS do not have direct ecological meaning but represent directions of maximum variation in community composition.

### Reading the Ordination

- **Proximity:** Sites close to each other have similar species composition
- **Distance:** Greater distances indicate greater compositional dissimilarity
- **Clusters:** Groups of points may indicate distinct community types
- **Gradients:** Smooth transitions may reflect environmental gradients

\vspace{0.5cm}

\newpage

# Methods

## Non-metric Multidimensional Scaling (NMDS)

NMDS is an ordination technique that attempts to represent the dissimilarity between samples in a low-dimensional space while preserving the rank order of dissimilarities (Kruskal, 1964). Unlike metric ordination methods (PCA, CA), NMDS:

1. **Does not assume linear relationships** between variables
2. **Preserves rank-order** of dissimilarities rather than exact distances
3. **Is robust to non-normal distributions** and zeros in the data
4. **Works with any dissimilarity measure**

### Analysis Parameters

- **Distance Metric:** `r toupper(params$distance)` dissimilarity
- **Dimensions:** `r params$k`
- **Number of sites:** `r nrow(scores(params$nmds_result, display = "sites"))`
- **Number of species:** `r nrow(scores(params$nmds_result, display = "species"))`
- **Random starts:** `r params$nmds_result$tries`
- **Convergence criterion:** Stress reduction

### Algorithm

The NMDS algorithm iteratively adjusts sample positions to minimize stress, which measures the departure from monotonicity in the relationship between observed dissimilarities and ordination distances. The algorithm:

1. Starts with a random (or PCA-based) configuration
2. Calculates distances between points in ordination space
3. Compares these with observed dissimilarities
4. Moves points to reduce stress
5. Repeats until convergence or maximum iterations reached

## Stress Interpretation Guidelines

The quality of NMDS ordination is assessed using the stress value (Clarke, 1993):

- **Stress < 0.05:** Excellent representation with no prospect of misinterpretation
- **Stress < 0.10:** Good ordination with no real risk of drawing false inferences  
- **Stress < 0.20:** Potentially useful, but interpret with caution
- **Stress ≥ 0.20:** Poor representation; points are close to being arbitrarily placed

## Goodness of Fit

The goodness of fit for each site indicates how well that sample is represented in the ordination. Values close to 1 indicate good representation, while lower values suggest the sample's position may not accurately reflect its true dissimilarity to other samples.

## Shepard Diagram

The Shepard plot displays the relationship between:

- **X-axis:** Observed dissimilarities (from the original distance matrix)
- **Y-axis:** Ordination distances (Euclidean distances in the reduced space)
- **Step line:** Monotonic relationship that NMDS attempts to achieve
- **Scatter:** Actual fit of each pairwise comparison

Points closer to the step line indicate better representation of those particular sample pairs.

\vspace{0.5cm}

\newpage

# Reproducibility

## Analysis Configuration

This section documents all parameters and settings used in the analysis to ensure full reproducibility.

### Dataset Information

- **Dataset Name:** `r params$dataset_name`
- **Number of Sites/Samples:** `r if(!is.null(params$n_sites)) params$n_sites else nrow(scores(params$nmds_result, display = "sites"))`
- **Number of Species/Variables:** `r if(!is.null(params$n_species)) params$n_species else nrow(scores(params$nmds_result, display = "species"))`
- **Analysis Date:** `r if(!is.null(params$analysis_date)) format(params$analysis_date, '%B %d, %Y') else format(Sys.Date(), '%B %d, %Y')`
- **Analysis Time:** `r if(!is.null(params$analysis_time)) params$analysis_time else format(Sys.time(), '%H:%M:%S')`

### NMDS Parameters

The following parameters were used for the NMDS ordination:

```{r repro_table}
repro_df <- data.frame(
  Parameter = c(
    "Distance/Dissimilarity Metric",
    "Number of Dimensions (k)",
    "Maximum Random Starts",
    "Autotransformation",
    "Permutations (stress test)",
    "Convergence Achieved",
    "Final Iterations",
    "Final Stress Value"
  ),
  Value = c(
    toupper(params$distance),
    as.character(params$k),
    as.character(if(!is.null(params$trymax)) params$trymax else params$nmds_result$tries),
    if(!is.null(params$autotransform)) ifelse(params$autotransform, "Yes", "No") else "No",
    as.character(if(!is.null(params$permutations)) params$permutations else "999"),
    if(params$nmds_result$converged) "✓ Yes" else "✗ No",
    as.character(params$nmds_result$iters),
    sprintf("%.4f", params$nmds_result$stress)
  ),
  Description = c(
    "Dissimilarity measure used to quantify differences between samples",
    "Number of ordination axes/dimensions",
    "Number of random starting configurations attempted",
    "Whether data were automatically transformed (e.g., Wisconsin double standardization)",
    "Number of permutations for statistical testing",
    "Whether the iterative algorithm reached a stable solution",
    "Number of iterations before convergence or termination",
    "Final stress value indicating quality of ordination"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(repro_df, 
             format = "latex", 
             booktabs = TRUE,
             caption = "Complete NMDS Analysis Configuration",
             longtable = TRUE,
             align = c("l", "l", "l")) %>%
  kable_styling(latex_options = c("HOLD_position", "repeat_header", "striped"),
                font_size = 9)
```

### Software Environment

```{r software_table}
software_df <- data.frame(
  Component = c(
    "Analysis Platform",
    "Platform Version",
    "R Version",
    "vegan Package",
    "Operating System"
  ),
  Version = c(
    "Ördin",
    if(!is.null(params$ordin_version)) paste0("v", params$ordin_version) else "v3.0",
    if(!is.null(params$r_version)) params$r_version else as.character(getRversion()),
    if(!is.null(params$vegan_version)) params$vegan_version else as.character(packageVersion("vegan")),
    Sys.info()["sysname"]
  ),
  stringsAsFactors = FALSE
)

knitr::kable(software_df, 
             format = "latex", 
             booktabs = TRUE,
             caption = "Software Environment",
             align = c("l", "l")) %>%
  kable_styling(latex_options = c("HOLD_position", "striped"))
```

### Reproducing This Analysis

To reproduce this exact analysis:

1. **Load your data** into R with the same structure:
   - Rows = Sites/Samples (`r if(!is.null(params$n_sites)) params$n_sites else nrow(scores(params$nmds_result, display = "sites"))` sites)
   - Columns = Species/Variables (`r if(!is.null(params$n_species)) params$n_species else nrow(scores(params$nmds_result, display = "species"))` species)

2. **Install required packages:**
   ```r
   install.packages("vegan")
   library(vegan)
   ```

3. **Run NMDS with identical parameters:**
   ```r
   nmds_result <- metaMDS(
     comm = your_data,           # Your community matrix
     distance = "`r params$distance`",
     k = `r params$k`,
     trymax = `r if(!is.null(params$trymax)) params$trymax else params$nmds_result$tries`,
     autotransform = `r if(!is.null(params$autotransform)) toupper(as.character(params$autotransform)) else "FALSE"`,
     trace = FALSE
   )
   ```

4. **Verify convergence:**
   ```r
   nmds_result$converged  # Should be TRUE
   nmds_result$stress     # Should be ≈ `r sprintf("%.4f", params$nmds_result$stress)`
   ```

**Note:** Due to random starting configurations, exact site coordinates may differ slightly between runs, but stress values and overall patterns should be nearly identical.

\vspace{0.5cm}

\newpage

# References

Clarke, K.R. (1993). Non-parametric multivariate analyses of changes in community structure. *Australian Journal of Ecology*, 18(1), 117-143. https://doi.org/10.1111/j.1442-9993.1993.tb00438.x

Kruskal, J.B. (1964). Multidimensional scaling by optimizing goodness of fit to a nonmetric hypothesis. *Psychometrika*, 29(1), 1-27. https://doi.org/10.1007/BF02289565

Minchin, P.R. (1987). An evaluation of the relative robustness of techniques for ecological ordination. *Vegetatio*, 69(1-3), 89-107. https://doi.org/10.1007/BF00038690

Oksanen, J., et al. (2024). vegan: Community Ecology Package. R package version 2.7-2. https://CRAN.R-project.org/package=vegan

\vspace{0.5cm}

# Software

This analysis was performed using:

- **Ördin v3.0** - Community Ecology Analysis Platform
- **R** version `r getRversion()`
- **vegan** package version `r packageVersion("vegan")`
- **Generated:** `r format(Sys.time(), '%Y-%m-%d %H:%M:%S %Z')`

\vspace{1cm}

---

**Report generated by Ördin v3.0**  
**Developed by:** Jimmy Moses  
**Email:** jmoses@pnguot.ac.pg  
**Institution:** Papua New Guinea University of Technology  

**Citation:** If you use Ördin for your research, please cite:  
Moses, J. (2025). Ördin: A cross-platform desktop application for community ecology analysis. Version 3.0.

\vspace{0.5cm}

**License:** This report and the Ördin software are provided for ecological research and education.
