---
title: "NMDS Ordination Analysis Report"
author: "Ördin v3.0"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  pdf_document:
    toc: false
    number_sections: false
geometry: margin=1in
header-includes:
  - \usepackage{xcolor}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhead[L]{NMDS Analysis Report}
  - \fancyhead[R]{\thepage}
  - \fancyfoot[C]{Generated by Ördin - Community Ecology Analysis Platform}
params:
  nmds_result: NULL
  distance: "bray"
  k: 2
  stress_interp: NULL
  dataset_name: "Community Data"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center")
library(vegan)
```

\vspace{0.5cm}

# Analysis Overview

**Analysis Type:** Non-metric Multidimensional Scaling (NMDS)  
**Dataset:** `r params$dataset_name`  
**Number of Sites:** `r nrow(scores(params$nmds_result, display = "sites"))`  
**Number of Species:** `r nrow(scores(params$nmds_result, display = "species"))`  
**Distance Metric:** `r toupper(params$distance)`  
**Dimensions (k):** `r params$k`

\vspace{0.5cm}

---

\vspace{0.5cm}

# Ordination Quality Assessment

\colorbox{yellow!20}{\parbox{\dimexpr\textwidth-2\fboxsep}{
\textbf{Stress: `r sprintf("%.4f", params$nmds_result$stress)`} \quad \textbf{[Grade: `r params$stress_interp$grade`]}

\textbf{`r params$stress_interp$message`}

`r params$stress_interp$detail`

\textbf{Recommendation:} `r params$stress_interp$recommendation`
}}

\vspace{0.3cm}

**Convergence:** `r if(params$nmds_result$converged) "✓ Solution converged" else "✗ Solution did not converge"`  
**Iterations:** `r params$nmds_result$iters`

\vspace{0.5cm}

# NMDS Ordination Plot

```{r nmds_plot, fig.width=8, fig.height=6, dpi=300}
# Create NMDS plot
plot(params$nmds_result, 
     type = "none", 
     main = "NMDS Ordination",
     cex.main = 1.4,
     col.main = "#2e8b57")

# Add sample points
points(params$nmds_result, 
       display = "sites", 
       pch = 21,
       bg = "#2e8b57", 
       cex = 2,
       col = "black",
       lwd = 1.5)

# Add stress annotation
stress_text <- sprintf("Stress = %.3f [%s]", 
                      params$nmds_result$stress,
                      params$stress_interp$grade)
mtext(stress_text, 
      side = 3, 
      line = 0.5, 
      col = params$stress_interp$color,
      font = 2, 
      adj = 0,
      cex = 0.9)

# Add grid for reference
grid(col = "#cccccc40", lty = 1)
```

\vspace{0.5cm}

# Ordination Statistics

```{r stats_table}
stats_df <- data.frame(
  Statistic = c("Stress", "Quality Grade", "Convergence", "Dimensions", 
                "Distance Metric", "Iterations", "Try Max"),
  Value = c(
    sprintf("%.4f", params$nmds_result$stress),
    params$stress_interp$grade,
    if(params$nmds_result$converged) "Converged" else "Not Converged",
    params$k,
    toupper(params$distance),
    params$nmds_result$iters,
    params$nmds_result$tries
  ),
  stringsAsFactors = FALSE
)

knitr::kable(stats_df, 
             format = "latex", 
             booktabs = TRUE,
             caption = "NMDS Analysis Statistics")
```

\newpage

# Site Scores

```{r site_scores}
site_scores <- as.data.frame(scores(params$nmds_result, display = "sites"))
site_scores$Sample <- rownames(site_scores)
site_scores <- site_scores[, c("Sample", colnames(site_scores)[1:(ncol(site_scores)-1)])]

knitr::kable(site_scores, 
             format = "latex", 
             booktabs = TRUE,
             digits = 4,
             caption = "NMDS Site Scores",
             longtable = TRUE)
```

\vspace{1cm}

# Methods

## Non-metric Multidimensional Scaling (NMDS)

NMDS is an ordination technique that attempts to represent the dissimilarity between samples in a low-dimensional space while preserving the rank order of dissimilarities. Unlike metric ordination methods, NMDS does not assume linear relationships between variables.

**Distance Metric:** `r toupper(params$distance)` dissimilarity was used to quantify differences between samples.

**Stress:** The stress value indicates how well the ordination represents the original dissimilarities. Lower stress values indicate better representation.

## Stress Interpretation Guidelines

The quality of NMDS ordination is assessed using the stress value, following guidelines by Clarke (1993):

- **Stress < 0.05:** Excellent representation with no prospect of misinterpretation
- **Stress < 0.10:** Good ordination with no real risk of drawing false inferences  
- **Stress < 0.20:** Potentially useful, but should be used with caution
- **Stress ≥ 0.20:** Poor representation; points are close to being arbitrarily placed

\vspace{0.5cm}

# References

Clarke, K.R. (1993). Non-parametric multivariate analyses of changes in community structure. *Australian Journal of Ecology*, 18(1), 117-143.

Oksanen, J., et al. (2024). vegan: Community Ecology Package. R package version 2.6-8.

\vspace{1cm}

---

**Report generated by Ördin v3.0**  
**Author:** Jimmy Moses (jmoses@pnguot.ac.pg)  
**Institution:** Papua New Guinea University of Technology  
**Date:** `r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`
