# Reproducibility

## Analysis Configuration

This section documents all parameters and settings used in the analysis to ensure full reproducibility.

### Dataset Information

- **Dataset Name:** `r params$dataset_name`
- **Number of Sites/Samples:** `r if(!is.null(params$n_sites)) params$n_sites else "Not specified"`
- **Number of Species/Variables:** `r if(!is.null(params$n_species)) params$n_species else "Not specified"`
- **Analysis Type:** `r if(!is.null(params$analysis_type)) params$analysis_type else "Not specified"`
- **Analysis Date:** `r if(!is.null(params$analysis_date)) format(params$analysis_date, '%B %d, %Y') else format(Sys.Date(), '%B %d, %Y')`
- **Analysis Time:** `r if(!is.null(params$analysis_time)) params$analysis_time else format(Sys.time(), '%H:%M:%S')`

### Analysis Parameters

```{r repro_params_table, eval=exists("repro_df", inherits=FALSE)}
# This chunk renders the analysis-specific parameters table
# The parent report must define 'repro_df' data frame with columns:
# - Parameter: Name of the parameter
# - Value: Value used in the analysis  
# - Description: What the parameter means

if (exists("repro_df", inherits = FALSE)) {
  knitr::kable(repro_df, 
               format = "latex", 
               booktabs = TRUE,
               caption = paste(params$analysis_type, "Configuration"),
               longtable = TRUE,
               align = c("l", "l", "l")) %>%
    kable_styling(latex_options = c("HOLD_position", "repeat_header", "striped"),
                  font_size = 9)
}
```

### Software Environment

```{r software_environment_table}
# Build software environment table
software_df <- data.frame(
  Component = c(
    "Analysis Platform",
    "Platform Version",
    "R Version",
    "Operating System"
  ),
  Version = c(
    "Ördin",
    if(!is.null(params$ordin_version)) paste0("v", params$ordin_version) else "v3.0",
    if(!is.null(params$r_version)) params$r_version else as.character(getRversion()),
    if(!is.null(params$os)) params$os else Sys.info()["sysname"]
  ),
  stringsAsFactors = FALSE
)

# Add key package versions if available
package_params <- c("vegan_version", "inext_version", "ggplot2_version", 
                   "rmarkdown_version", "knitr_version")

for (pkg_param in package_params) {
  if (!is.null(params[[pkg_param]])) {
    pkg_name <- gsub("_version", "", pkg_param)
    software_df <- rbind(software_df, data.frame(
      Component = paste0(pkg_name, " package"),
      Version = params[[pkg_param]],
      stringsAsFactors = FALSE
    ))
  }
}

knitr::kable(software_df, 
             format = "latex", 
             booktabs = TRUE,
             caption = "Software Environment",
             align = c("l", "l")) %>%
  kable_styling(latex_options = c("HOLD_position", "striped"))
```

### Reproducing This Analysis

To reproduce this exact analysis:

```{r repro_instructions, eval=exists("repro_code", inherits=FALSE), results='asis'}
# This chunk renders reproduction code if provided
# The parent report must define 'repro_code' character string

if (exists("repro_code", inherits = FALSE)) {
  cat(repro_code)
} else {
  # Default generic instructions
  cat(paste0(
    "1. **Load your data** with the same structure:\\n",
    "   - Rows = Sites/Samples (", 
    if(!is.null(params$n_sites)) params$n_sites else "N", " sites)\\n",
    "   - Columns = Species/Variables (", 
    if(!is.null(params$n_species)) params$n_species else "M", " species)\\n\\n",
    "2. **Install required packages** as shown in the Software Environment table above.\\n\\n",
    "3. **Run the analysis** using the parameters documented in the Analysis Parameters table.\\n\\n",
    "4. **Verify results** match the values reported in this document.\\n"
  ))
}
```

**Note:** Due to random number generation (if applicable), exact numerical values may differ slightly between runs, but overall patterns and statistical conclusions should be identical. For exact reproducibility, set a random seed before running the analysis.

\vspace{0.5cm}

### Session Information

For complete transparency, the following packages were loaded during this analysis:

```{r session_info, eval=FALSE}
# Session info can be captured with:
sessionInfo()
```

**Important:** Save this report along with your original data to maintain a complete audit trail of your analysis.
